---
output: html_document
---

## Dealing with Confounders

### Introduction

In the previous chapter we normalized for library size, effectively removing it as a confounder. Now we will consider removing other less well-defined confounders from our data. Technical confounders (aka batch effects) can arise from difference in reagents, isolation methods, the lab/experimenter who performed the experiment, even which day or time of the day the experiment was performed. Accounting for technical confounders, and batch effects particularly, is a large topic that also involves principles of experimental design. Here we address approaches that can be taken to account for confounders when the experimental design is appropriate.

Fundamentally, accounting for technical confounders involves identifying and, ideally, removing sources of variation in the expression data that are not related to (i.e. are confounding) the biological signal of interest. Various approaches exist, some of which use spike-in or housekeeping genes, and some of which use endogenous genes.

The use of spike-ins as control genes is appealing, since the same amount of ERCC (or other) spike-in was added to each cell in our experiment. In principle, all the variablity we observe for these genes is due to technical noise; whereas endogenous genes are affected by both technical noise and biological variability. Technical noise can be removed by fitting a model to the spike-ins and "substracting" this from the endogenous genes. There are several methods available based on this premise (eg. [BASiCS](https://github.com/catavallejos/BASiCS), [scLVM](https://github.com/PMBio/scLVM), [RUVg](http://bioconductor.org/packages/release/bioc/html/RUVSeq.html)); each using different noise models and different fitting procedures. Alternatively, one can identify genes which exhibit significant variation beyond technical noise (eg. Distance to median, [Highly variable genes](http://www.nature.com/nmeth/journal/v10/n11/full/nmeth.2645.html)). However, there are issues with the use of spike-ins for normalisation (particularly ERCCs, derived from bacterial sequences), including that their variability can, for various reasons, actually be *higher* than that of endogenous genes.

Given the issues with using spike-ins, better results can often be obtained by using endogenous genes instead. Where we have a large number of endogenous genes that, on average, do not vary systematically between cells and where we expect technical effects to affect a large number of genes (a very common and reasonable assumption), then such methods (for example, the RUVs method) can perform well. 

There are two scenarios in scRNA-seq dataset integration. In the first scenario, cell composition is expected to be the same, and methods developed for bulk RNA-seq (e.g. ComBat) exhibit good performance. This is often true for biological replicates of the same experiment; this is also true for batches in `tung` dataset. In the second scenario, the overlap between the datasets is partial - e.g. if datasets represent healthy and diseased tissue, which differ in cell type composition substantially. In this case, mutual nearest neighbor (MNN)-based methods tend to perform much better. We will look at these

Here, we will perform batch correction using two methods - `ComBat`, based on empirical Bayesian framework, and `fastMNN`, which is a MNN-based method from the package `batchelor`.

### Load and Normalize the Tung Dataset

```{r remove-conf0, echo=FALSE}
library(knitr)
opts_chunk$set(cache= TRUE, fig.align = 'center')
```

```{r remove-conf1, message=FALSE, warning=FALSE}
library(scRNA.seq.funcs)
library(scater)
library(scran)
library(sva)
library(batchelor)
library(kBET)
set.seed(1234567)
```

Let's read in the pre-processed dataset and normalize it using `logNormCounts` from `scran` package. In `umi.qc` object, a new assay named `logcounts` will appear, in addition to the previously present `counts` and `logcounts_raw`: 

```{r remove-conf2}
umi    <- readRDS("data/tung/umi.rds")
umi.qc <- umi[! rowData(umi)$discard, ! colData(umi)$discard]
qclust <- quickCluster(umi.qc, min.size = 30)
umi.qc <- computeSumFactors(umi.qc, clusters = qclust)
umi.qc <- logNormCounts(umi.qc)
```

### Combat

If you have an experiment with a balanced design, `ComBat` can be used to eliminate batch effects while preserving biological effects by specifying the biological effects using the `mod` parameter. However the `Tung` data contains multiple experimental replicates rather than a balanced design so using `mod1` to preserve biological variability will result in an error. 

```{r remove-conf5}
assay(umi.qc, "combat") <- ComBat(logcounts(umi.qc),batch = umi.qc$replicate)
```

__Exercise 1__

Perform `ComBat` correction accounting for total features as a co-variate. Store the corrected matrix in the `combat_tf` slot.

<details><summary>Answer</summary>

```{r remove-conf6}
assay(umi.qc, "combat_tf") <- ComBat(logcounts(umi.qc),batch = umi.qc$detected)
```

</details>

### mnnCorrect (batchelor)

MNN-based normalization is implemented as `fastMNN` function.

```{r remove-conf7}
mnn_out <- fastMNN(umi.qc,batch = umi.qc$replicate)
assay(umi.qc, "mnn") <- assay(mnn_out,'reconstructed')
```


### Evaluation and Comparison of Batch-removal Approaches

A key question when considering the different methods for removing confounders is how to quantitatively determine which one is the most effective. The main reason why comparisons are challenging is because it is often difficult to know what corresponds to technical counfounders and what is interesting biological variability. Here, we consider three different metrics which are all reasonable based on our knowledge of the experimental design. Depending on the biological question that you wish to address, it is important to choose a metric that allows you to evaluate the confounders that are likely to be the biggest concern for the given situation.

#### Effectiveness 1

We evaluate the effectiveness of the normalization by inspecting the PCA plot where colour corresponds the technical replicates and shape corresponds to different biological samples (individuals). Separation of biological samples and interspersed batches indicates that technical variation has been removed. We always use log2-cpm normalized data to match the assumptions of PCA.

```{r remove-conf11}
for(n in assayNames(umi.qc)) {
    tmp <- runPCA(umi.qc, exprs_values = n, ncomponents = 20)
    
    print(
        plotPCA(
            tmp,
            colour_by = "batch",
            size_by = "detected",
            shape_by = "individual"
        ) +
        ggtitle(n)
    )
}
```

#### Effectiveness 2

We can also examine the effectiveness of correction using the relative log expression (RLE) across cells to confirm technical noise has been removed from the dataset. Note RLE only evaluates whether the number of genes higher and lower than average are equal for each cell - i.e. systemic technical effects. Random technical noise between batches may not be detected by RLE.

```{r remove-conf12}

res <- list()
for(n in assayNames(umi.qc)) {
	res[[n]] <- suppressWarnings(calc_cell_RLE(assay(umi.qc, n)))
}
par(mar=c(6,4,1,1))
boxplot(res, las=2)
```

#### Effectiveness 3

Another method to check the efficacy of batch-effect correction is to consider the intermingling of points from different batches in local subsamples of the data. If there are no batch-effects then proportion of cells from each batch in any local region should be equal to the global proportion of cells in each batch. 

`kBET` [@Buttner2017-ds] takes `kNN` networks around random cells and tests the number of cells from each batch against a binomial distribution. The rejection rate of these tests indicates the severity of batch-effects still present in the data (high rejection rate = strong batch effects). `kBET` assumes each batch contains the same complement of biological groups, thus it can only be applied to the entire dataset if a perfectly balanced design has been used. However, `kBET` can also be applied to replicate-data if it is applied to each biological group separately. In the case of the Tung data, we will apply `kBET` to each individual independently to check for residual batch effects. However, this method will not identify residual batch-effects which are confounded with biological conditions. In addition, `kBET` does not determine if biological signal has been preserved. 

```{r remove-conf13, message = FALSE, warning = FALSE}
compare_kBET_results <- function(sce){
    sce <- umi.qc
    indiv <- unique(as.character(sce$individual))
    norms <- assayNames(sce) # Get all normalizations
    results <- list()
    for (i in indiv){ 
        for (j in norms){
            tmp <- kBET(
                df = t(assay(sce[,sce$individual== i], j)), 
                batch = sce$batch[sce$individual==i], 
                heuristic = TRUE, 
                verbose = FALSE, 
                addTest = FALSE, 
                plot = FALSE)
            results[[i]][[j]] <- tmp$summary$kBET.observed[1]
        }
    }
    return(do.call(rbind.data.frame, results))
}

eff_debatching <- compare_kBET_results(umi.qc)
eff_debatching
```

Finally, let's visualize the output of kBET calculation:

```{r remove-conf14, message = FALSE}
library("reshape2")
library("RColorBrewer")
# Plot results
dod <- melt(as.matrix(eff_debatching),  value.name = "kBET")
colnames(dod)[1:2] <- c("Normalisation", "Individual")

colorset <- c('gray', brewer.pal(n = 9, "RdYlBu"))

ggplot(dod, aes(Normalisation, Individual, fill=kBET)) +  
    geom_tile() +
    scale_fill_gradient2(
        na.value = "gray",
        low = colorset[2],
        mid=colorset[6],
        high = colorset[10],
        midpoint = 0.5, limit = c(0,1)) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) + 
    theme(
        axis.text.x = element_text(
            angle = 45, 
            vjust = 1, 
            size = 12, 
            hjust = 1
        )
    ) + 
    ggtitle("Effect of batch regression methods per individual")
```

### sessionInfo()

<details><summary>View session info</summary>
```{r echo=FALSE}
sessionInfo()
```
</details>

